Making ice sheet models \\ scale properly

### 2023-01-22 talk/slides.tex
last section still needs to be written

\section{Stokes solver scaling}

@@

\begin{itemize}
\item from now on assume our ice sheet model is \emph{not} shallow
\item whether time steps are explicit or implicit, a Stokes problem is solved at each step
\item for a given icy domain $\Lambda \subset \RR^3$:
\begin{align*}
- \nabla \cdot \left(2 \nu_\eps(D\bu)\, D\bu\right) + \nabla p - \rhoi \mathbf{g} &= \bzero && \text{in $\Lambda$} \\
\nabla \cdot \bu &= 0 && \text{''} \\
\btau_b - \bbf(\bu|_b) &= \bzero && \text{on $\Gamma_b$} \\
\bu|_b \cdot \bn_b &= 0 && \text{''} \\
\left(2 \nu_\eps(D\bu) D\bu - pI\right) \bn &= \bzero && \text{on $\Gamma_s$}
\end{align*}
\item how fast is the numerical solution process?
\end{itemize}
\end{frame}

\begin{frame}{PDE solver complexity}

FIXME explain for Poisson
\end{frame}

\begin{frame}{a scalable Stokes solver for ice sheets}

FIXME Isaac et al.~2015
\end{frame}

\begin{frame}{a scalable Stokes solver for ice sheets 2}

FIXME table from Isaac, and estimate power $\alpha$
\end{frame}

\section{simplified performance analysis}

@@

FIXME asymptotic estimates of algorithmic scaling, measured by floating point operations per model year, for map-plane (2D) time-stepping numerical ice sheet simulations, in the high resolution limit where $\Delta x\to 0$ and $m\to\infty$
\end{frame}

\begin{frame}{ice sheet model performance: table}

\begin{tabular}{llll}
\emph{time-stepping} & \emph{dynamics} & \emph{flops per model year} \\ \hline
\,explicit & SIA    & $\oo{\frac{D\, L^2}{\Delta x^4}} = \oo{\frac{D\, m^2}{L^2}}$ \\
explicit & Stokes (\emph{optimistic}) & $\oo{\frac{U L^{2+2\alpha}}{\Delta x^{3+2\alpha}}} = \oo{\frac{U m^{1.5+\alpha}}{L}}$ \\
explicit & Stokes (\emph{pessimistic})  & $\oo{\frac{D\, L^{2+2\alpha}}{\Delta x^{4+2\alpha}}} = \oo{\frac{D\,m^{2+\alpha}}{L^2}}$
\end{tabular}

FIXME table from Bueler (2022)
\end{frame}

\begin{frame}{complexity for implicit solves}

FIXME $\beta$, $\gamma$
\end{frame}

\begin{frame}{ice sheet model performance: table cont.}

\begin{tabular}{llll}
\emph{time-stepping} & \emph{dynamics} & \emph{flops per model year} \\ \hline
\,explicit & SIA    & $\oo{\frac{D\, L^2}{\Delta x^4}} = \oo{\frac{D\, m^2}{L^2}}$ \\
explicit & Stokes (\emph{optimistic}) & $\oo{\frac{U L^{2+2\alpha}}{\Delta x^{3+2\alpha}}} = \oo{\frac{U m^{1.5+\alpha}}{L}}$ \\
explicit & Stokes (\emph{pessimistic})  & $\oo{\frac{D\, L^{2+2\alpha}}{\Delta x^{4+2\alpha}}} = \oo{\frac{D\,m^{2+\alpha}}{L^2}}$ \\
implicit & SIA    & $\oo{\frac{q\, L^{2+2\beta}}{\Delta x^{2+2\beta}}} = \oo{q\, m^{1+\beta}}$ \\
implicit & Stokes & $\oo{\frac{q\, L^{2+2\gamma}}{\Delta x^{2+2\gamma}}} = \oo{q\, m^{1+\gamma}}$
\end{tabular}

FIXME table from Bueler (2022)
\end{frame}

@@

\begin{itemize}
\item ice sheet models solve a multi-scale, irregular-data problem with hard-to-observe boundary conditions
   \begin{itemize}
   \item[$\circ$] there are \alert{no easy or magic techniques} for performance
   \end{itemize}
\item<2-> current-technology ice sheet models mostly use \alert{explicit} time stepping, \alert{non-optimal} stress-balance solvers, and \alert{shallow} assumptions
   \begin{itemize}
   \item[$\circ$] progress will be made in all of these areas
   \end{itemize}
\item<3-> \alert{coming soon} from current research:
   \begin{enumerate}
   \item[1.] machine learning emulators (Jouvet et al.~2021)
   \item[2.] semi-implicit, semi-coupled time stepping (L{\"o}fgren et al.~2022)
   \item[3.] scalable Stokes solvers (Isaac et al.~2015)   
   \end{enumerate}
\item<4-> scalable solvers for the implicit-step problem require advances in \alert{multilevel solutions of non-local variational inequalities}
\end{itemize}
\end{frame}

>>>>

\section{stress balance solver scaling}

@@

\begin{itemize}
\item assume an ice sheet model which is \emph{not} shallow
\item whether explicit or implicit, solves a Stokes problem at each step:
\begin{align*}
- \nabla \cdot \left(2 \nu_\eps(D\bu)\, D\bu\right) + \nabla p - \rhoi \mathbf{g} &= \bzero && \text{in $\Lambda$} \\
\nabla \cdot \bu &= 0 && \text{''} \\
\btau_b - \bbf(\bu|_b) &= \bzero && \text{on $\Gamma_b$} \\
\bu|_b \cdot \bn_b &= 0 && \text{''} \\
\left(2 \nu_\eps(D\bu) D\bu - pI\right) \bn &= \bzero && \text{on $\Gamma_s$}
\end{align*}
\item this problem is the \alert{stress balance} (conservation of momentum) which determines velocity $\bu$ and pressure $p$
\item how fast is the numerical solution process?
\end{itemize}
\end{frame}

\begin{frame}{PDE solver algorithmic scaling}

\begin{itemize}
\item as intro to scaling, consider the 3D Poisson equation:
    $$-\grad^2 u = f \text{ in $\Omega$}, \quad u=0 \text{ on $\partial\Omega$}$$
\item discretization generates a linear system $A\bu = \bb$
\item (data size $m$) = \alert{number of unknowns}
    \begin{itemize}
    \item[$\circ$] for low-order discretizations, $m$ = (number of nodes in the grid)
    \item[$\circ$] for cell size $h$, $m \sim h^{-3}$ in 3D
    \end{itemize}
\item solver choice determines asymptotic growth of computational expense (complexity)
\item \alert{algorithmic scaling} of flops as $m\to\infty$:
    \begin{itemize}
    \item[$\circ$] $O(m^3)$ for direct linear algebra, ignoring matrix structure
    \item[$\circ$] $\approx O(m^2)$ for sparsity-exploiting direct linear algebra
    \item[$\circ$] $O(m^1)$, \alert{optimal}, for \alert{multigrid} solvers (below)
    \end{itemize}
\end{itemize}

\medskip
\begin{center}
\includegraphics[width=0.5\textwidth]{images/multigrid-grids.png}
\end{center}
\end{frame}

\begin{frame}{scalable stress balance solvers for ice sheets}

\begin{itemize}
\item ice sheet stress balance solvers can be analyzed similarly
\item $m$ = (number of velocity and pressure unknowns)
    \begin{itemize}
    \item[$\circ$] indicate scaling by ``$O(m^{1+\alpha})$'', with $\alpha=0$ optimal
    \end{itemize}
\item \alert{near-optimal solvers} already exist: \hfill \emph{good news!}
    \begin{itemize}
    \item[$\circ$] $\alpha=0.08$ for Isaac et al.~(2015) Stokes solver
        \begin{itemize}
        \item[$\vartriangleright$] unstructured quadrilateral/tetrahedral mesh, $Q_k\times Q_{k-2}$ stable elements, Schur-preconditioned Newton-Krylov, column-oriented algebraic multigrid (AMG) preconditioner for $(\bu,\bu)$ block
        \end{itemize}

\begin{center}
\includegraphics[width=0.6\textwidth]{images/isaac-antarctica.png}
\end{center}
    \item[$\circ$] $\alpha=0.05$ for Tuminaro et al~(2016) 1st-order (shallow) AMG solver
    \item[$\circ$] similar for Brown et al~(2013) GMG solver
    \end{itemize}
\end{itemize}
\end{frame}

\section{ice sheet model performance analysis}

@@

\begin{itemize}
\item for a time-stepping model our metric is \alert{flops per model year}
\item $m$ = (\# of surface elevation, velocity, and pressure unknowns)
\item domain $\Omega \subset \RR^2$ (map-plane) has characteristic width $L$
\item map-plane (2D) mesh cells have characteristic diameter $h$:
  $$m \sim \frac{L^2}{h^2}  \quad \iff \quad h \sim \frac{L}{m^{1/2}}$$
\item recall mass-continuity equation schemata:
{\footnotesize
\begin{align*}
\text{advective schema:} && \frac{\partial H}{\partial t} + \nabla \cdot \left(\bU H\right) &= a \\
\text{diffusion schema:} && \frac{\partial H}{\partial t} - \nabla \cdot \left(D \grad s\right) &= a
\end{align*}
}
\end{itemize}
\end{frame}

\begin{frame}{ice sheet model performance}

\begin{tabular}{llll}
\emph{time-stepping} &  & \emph{flops per model year} \\ \hline
\\
explicit & SIA    & $\oo{\frac{D\, L^2}{h^4}} = \oo{\frac{D}{L^2} m^2}$ \\
\\
explicit ({\footnotesize \emph{advective}}) & Stokes & $\oo{\frac{U L^{2+2\alpha}}{h^{3+2\alpha}}} = \oo{\frac{U}{L} m^{1.5+\alpha}}$ \\
\\
\phantom{explicit} ({\footnotesize \emph{diffusive}})  & Stokes & $\oo{\frac{D\, L^{2+2\alpha}}{h^{4+2\alpha}}} = \oo{\frac{D}{L^2} m^{2+\alpha}}$
\end{tabular}
\end{frame}

\begin{frame}{implicit time-stepping for ice sheet models}

\begin{itemize}
\item main point: \alert{explicit time-stepping stability restrictions imply many stress-balance solves}
    \begin{itemize}
    \item[$\circ$] $\alpha$ for stress balance solver important
    \item[$\circ$] however, even optimality ($\alpha=0$) cannot fix low model performance
    \end{itemize}
\item switch to \alert{implicit time-stepping} for unconditional stability?
    \begin{itemize}
    \item[$\circ$] each step is a NCP-coupled-to-Stokes problem
        \begin{itemize}
        \item[$\vartriangleright$] or SIA NCP
        \item[$\vartriangleright$] in any case, a \alert{free-boundary problem}
        \item[$\vartriangleright$] next slide parameterizes expense of these coupled, implicit-step solves as $O(m^{1+\beta})$; optimal is $\beta=0$
        \end{itemize}
    \item[$\circ$] need $q$ model updates per year to integrate climate influences
    \end{itemize}

\bigskip
\item<2> \alert{existing} implicit time-stepping ice sheet models:
    \begin{itemize}
    \item[$\circ$] Bueler (2016) SIA NCP solver scales badly ($\beta=0.8$)
    \item[$\circ$] no other fully-implicit NCP solvers known?
    \end{itemize}
\end{itemize}
\end{frame}

\begin{frame}{ice sheet model performance 2}

\begin{tabular}{llll}
\emph{time-stepping} &  & \emph{flops per model year} \\ \hline
\\
explicit & SIA    & $\oo{\frac{D\, L^2}{h^4}} = \oo{\frac{D}{L^2} m^{\,2}}$ \\
\\
explicit \, {\footnotesize \emph{advective}} & Stokes & $\oo{\frac{U L^{2+2\alpha}}{h^{3+2\alpha}}} = \oo{\frac{U}{L} m^{1.5+\alpha}}$ \\
\\
\phantom{explicit} \, {\footnotesize \emph{diffusive}}  & Stokes & $\oo{\frac{D\, L^{2+2\alpha}}{h^{4+2\alpha}}} = \oo{\frac{D}{L^2} m^{2+\alpha}}$ \\
\\
implicit & \emph{either} & $\oo{\frac{q\, L^{2+2\beta}}{h^{2+2\beta}}} = \oo{q\, m^{1+\beta}}$
\end{tabular}
\end{frame}

@@

\begin{itemize}
\item ice sheet models solve a multi-scale, irregular-data problem with hard-to-observe boundary conditions
   \begin{itemize}
   \item[$\circ$] there are \alert{no easy or magic techniques} for performance
   \end{itemize}
\item<2-> current-technology ice sheet models mostly use \alert{explicit} time stepping, \alert{non-optimal} stress-balance solvers, and \alert{shallow} assumptions
   \begin{itemize}
   \item[$\circ$] progress will be made in all of these areas
   \end{itemize}
\item<3-> \alert{coming soon} from current research:
   \begin{enumerate}
   \item[1.] machine learning emulators (Jouvet et al.~2021)
   \item[2.] semi-implicit, semi-coupled time stepping (L{\"o}fgren et al.~2022)
   \item[3.] scalable Stokes solvers (Isaac et al.~2015)   
   \end{enumerate}
\item<4-> scalable solvers for implicit-step models require advances in \alert{multilevel solutions of non-local variational inequalities}
\end{itemize}
\end{frame}
